---
title: "Fixing the problems with lapply"
author: "Jonas Moss"
date: '2018-04-28'
slug: lapply2
tags:
- R
categories:
- programming
---

The [R](https://www.r-project.org/) language is a functional language. As most 
functional languages, it has a built-in function to handle the problem of 
applying a function to each element of a list. Such functions are usually 
called [*map*](https://en.wikipedia.org/wiki/Map_(higher-order_function)).

In R, *map* is called *lapply*. This article is about three problems with 
lapply and a method to solve two of them. 

The first problem is already alluded to. The name *lapply* is wrong, the name 
*map* is right. To make matters worse 'lapply' (and the rest of the functions
in this family, including *apply*) makes it impossible to rename 'do.call' to 
its proper name, which is [apply](http://clhs.lisp.se/Body/f_apply.htm). I 
won't be able to solve this problem, as the apply family is nailed tightly to 
`R`. But the name is wrong anyhow.

There are two more substantive problems though.

1. lapply doesn't respect side effects,
2. lapply doesn't preserve the call structure.

# Lapply doesn't respect side effects
Here's some code and some output.
```{r lapply-side-effects-good}
x = 0

f = function(y) {
  envir = parent.frame()
  assign("x", x + 1, envir = envir)
  y + x
}

list(f(1), f(2), f(3))
```

So `f` takes an argument `y`, augments `x` by 1 in its parent environment, 
and returns `x + y`. Easy to understand, and the list contains the expected 
elements. Now let's do the same with `lapply`:

```{r lapply-side-effects-bad}
x = 0

lapply(1:3, f)
```

Curiously, the `lapply` forgets to increase `x` by one each iteration! The 
explanation for this is pretty simple. When `lapply` is called, it evaluates 
all the function calls in its own environment. A 'fix' of the problem above is
to modify `f` in the following way:

```{r lapply-side-effects-better}
x = 0

f_fix = function(y) {
  envir = parent.frame(n = 2)
  assign("x", x + 1, envir = envir)
  y + x
}

lapply(1:3, f_fix)
```

But this is hardly a fix. It's a hack. Moreover, this function won't always work
when you call `f` repeatedly on a list's elements. Now I embed the `f` functions
inside `g` functions. According to the definition of `f`, only the `x` inside 
the environment of `g` should be changed. But `f_fix` doesn't fulfil this 
requirement.
 
```{r lapply-side-effects-worse}
x = 0

g = function() {
  x = 0
  list(f(1), f(2), f(3))
}

g_fix = function() {
  x = 0
  list(f_fix(1), f_fix(2), f_fix(3))
}

# Call for side-effects
tmp = g()
y = x
tmp = g_fix()
print(c(y, x))
```

That `x` should be 0, not 3. This happens due to our "fix" of `f`, which assigns
`x` two steps up the frame stack.

## Is this a problem?

Yes, this is a problem. I even guess that plenty of people have encountered this
problem without being aware of its cause. They would've rewritten the code,
avoided the `lapply`, and arrived at some code that did what they intended for
it to do. 

Furthermore, the behaviour can create tiny, opaque bugs that are really 
difficult to weed out. This is some behaviour we don't want from the basic
built-ins of a programming language.

## How it should be handled

`lapply` should 


Is this possible or reasonable? Yes. Just think about `list`. When list is 
called with `f(1)` etc., as in `list(f(1), f(2), f(3))`, nothing problematic 
happens. This is probably because `list` doesn't create its own evaluated 
environment, so that `x` is updated as it should be. 

# lapply doesn't preserve calls and arguments

```{r preserve-arguments}


f = function(x) substitute(x)

x = 1
y = 2
z = 3

list(f(x), f(y), f(z))
lapply(alist(x, y, z), f)
```

# A solution
Before I present a solution.

Here is my solution proposal. 
```{r solution-proposal}
lapply2 = function(X, FUN, ...) {
  FUN = substitute(FUN)
  dots = as.list(substitute((...))[-1])
  envir = parent.frame()
  lapply(X, function(x) eval(as.call(c(FUN, x, dots)), envir = envir))
}
```
