---
title: "Comments on Nordstat 2018"
author: "Jonas Moss"
date: '2018-07-01'
slug: nordstat-2018-comments
categories:
- statistics
---

## van der Vaart's talk
van der Vaart's talk was the highlight of the conference. 

## Fréchet means and smearyness



```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
Q = function(call, quote = TRUE, .force = TRUE, .enclos = NULL) {

  call = if(quote) substitute(call) else call
  name = deparse(call[[1]])
  object = getFunction(name)

  if(!is.null(.enclos)) environment(object) = .enclos

  args = as.list(match.call(object, call)[-1])

  if(.force) args = lapply(args, eval, envir = parent.frame(n = 3))

  for(arg in names(args)) formals(object)[[arg]] = args[[arg]]

  if(!is.null(.enclos)) environment(object) = .enclos

  object
}
```

I don't use partial function application enough. Say that I want to plot four
different curves in the same window. The curves will be defined by `f(x)` for 
some `f`, and most of them will share the `x`.

```{r}
x = seq(-1, 5, by = 0.01)
y = seq(-1, 2, by = 0.01)
plot(x, dnorm(x, mean = 1, sd = 2), type = "l")
lines(x, dnorm(x, mean = 1, sd = 1))
lines(x, dnorm(x, mean = 2, sd = 2))
points(y, dnorm(y, mean = 1, sd = 2))
```

But I'm not done rigth away. I want to add options, such as `lty`, `lwd`, `col`
and `bty` to these functions. What's more, I'll most likely use some time 
experimenting to find the best options for `x`, `y` and all these parameters.
Moreover, if I'm wrong about `sd = 2`, so that `sd = 3` holds instead, I must 
change every single `sd = 2` option.

Wouldn't it be nice to have a simple function to take care of these things? 
What if you could write, say,

```{r, eval = FALSE}
x = seq(-1, 5, by = 0.01)
dnorm2 = Q(dnorm(x = x, mean = 1, sd = 2))
```

to obtain a new dnorm that behaves exactly the same as if the original `dnorm`
in the `stats` package was defined with these default arguments? With the 
unreferced arguments, in this case `log = TRUE`, untouched?

## Defintion of the function

Here's the function definition. It's without error handling, as it makes it
easier to read. For a better function error-handling, see this gist. The gist
function includes some more options as well. 

```{r, eval = FALSE}
Q = function(call, quote = TRUE) {

  call   = if(quote) substitute(expr = call) else call
  name   = deparse(expr = call[[1]])
  object = match.fun(FUN = name)
  args   = lapply(X = as.list(match.call(object, call)[-1]), 
                  FUN = eval, 
                  envir = parent.frame(n = 3))

  for(arg in names(args)) formals(object)[[arg]] = args[[arg]]
  object
}
```

So what does this function do? First it identifies the function name from the 
supplied call (in `deparse(call[[1]])`) and fetches the function object 
associated with the call. Then it evaluates all the supplied arguments in the 
calling environment, matches them to the formals of the associated function 
object, and returns a copy of the associated function object with the formals 
changed.



## When default arguments are silly 

### Histograms
The default arguments in the histogram function `graphics::hist` are silly. 
First, it default to show a histogram with frequency counts on the y-axis, 
something I never want. The histogram is a density estimator (even a 
universally consisten one), and I want it to show a density. 
Hence `freq = FALSE` is the sensible default argument. Second, the default
choice of bin counts ("Sturges") is horrible, and is well known to be horrible. 
It uses far too few bins. I would have prefered a methods based on AIC, but it's
not implemented, so I'll go with Freedman-Diaconis, or the option 
`breaks = "FD"`. Since `hist` is an S3-generic, I will have to manipulate the 
`default` option. Thus I redefine the hist.default to:

```{r}
hist.default = Q(hist.default(freq = FALSE, breaks = "FD"))
``` 

(*Note:* This kind of practice isn't dangerous, and can't break code in other 
packages, as the generic is redefined in your current workspace. It can make 
it harder to copy-paste isolated parts of code, but that's very often a 
problem anyway.)

### Plotting
When I plot something, I usually want the plot to be a line plot.  

```{r}
plot.default = Q(plot.default(type = "l"))
``` 

Changing plotting parameters in this way is not very easy though, as `Q` can't 
work its way around `...` arguments. This creates problems when the function
you manipulate doesn't have your desired parameter as an argument, but rather in
`...`. An example of this is the `bty` parameter in `plot` (I always change this
to `"l"`). These problems reduce the usefulness of the Q function. 

### Using `lapply`
One of the main uses of partial function application lies in the use of `lapply`.
Here's an example,

## Problems and issues

