---
title: "Handling side effects with the .H function"
author: "Jonas Moss"
date: '2018-08-21'
slug: H-function
tags:
- R
- non-standard evaluation
categories:
- programming
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
.H = function(expr) {
  f = function() {}
  body(f) = substitute(expr)
  f()
}
```

### The Problem of Side Effects
When I am writing $\mathtt{R}$ code, I often do stuff in the body of my script that creates undesired side effects. 

```{r, eval = TRUE}
# n and x constants I wish to use later
n = 100
x = pi^2/6

# lots of code
# ...

# I suddenly wish to plot something
n = 1:1000
x = 0.1
plot(n, pnorm(sqrt(n)*x), type = "l")
```

Notice that `n` and `x` have been rewritten. So what should I do if I find myself in need of their original values?  Then I will have to hunt down where I've used `n` and `x` in my script and either change the variable names or put the entire paragraph in a function.

It's probably smarter to put it all in a function, as the new variable names (e.g. `n2` and `x2`) could also have important functions in the script. Since I obviously wish my function to be anonymous, my script will look like this:
```{r, eval = FALSE}
funtion() {
  n = 1:1000
  x = 0.1
  plot(n, pnorm(sqrt(n)*x)
}()
#
```

This code isn't beautiful at all.  It's also cryptic, as it doesn't convey the *purpose* of the function. 

### The Solution
To solve this problem I've made the *hide* function:

```{r, eval = FALSE} 
.H = function(expr) rlang::new_function(alist(), substitute(expr))()
```

Here "H" is for hide, and the dot in front of it makes the function hidden, which prevents it from cluttering your environment. 

Here's how `.H` works: It creates a function new function and executes it. The argument list of the new function is empty, this is why `alist()` appears. The body of the function is the argument passed to `.H`, that is the purpose of `substitute(expr)`. The package `rlang` is great for these sort of things, but it's easy to write `.H` without it. The following function is equivalent to `.H`:

```{r, eval = FALSE} 
.H2 = function(expr) {
  f = function() {}
  body(f) = substitute(expr)
  f()
}
```

I prefer the version using `rlang` since it is a one-liner.

With this function in hand, I can write

```{r, eval = TRUE}
.H({
  n = 1:1000
  x = 0.1
  plot(n, pnorm(sqrt(n)*x), type = "l")
})
```

### Conclusion
With the function `.H` I have reached my following three goals:

1. Plotting without side effects,
2. relatively pretty code,
3. code that is easy to understand, as `.H` is only used to avoid side effects of code that would usually be found in the body of the script.
