---
title: "Easy lambdas"
author: "Jonas Moss"
date: '2018-08-23'
slug: easy-lambda
tags:
- R
- non-standard evaluation
categories:
- programming
---

`R` already has anonymous functions. Just write `function(x) do_someting` and 
you're done. This post is for all of us who are not satisified with this 
solution. For one, it's exhausting to write to long `function(x)` instead of, 
say, `L(x)`, but that's not all of it! 

Take the following example, a calculation of the maximum likelihood estimate of 
a gamma distribution:

```{r, eval = TRUE}
set.seed(313)
x = rgamma(100, 2, 4)
nlm(f = function(p) -mean(dgamma(x, shape = p[1], rate = p[2], log = TRUE)), 
    p = c(1, 1))$estimate
```
Here the hassle is to keep track of the indices of `p`, which is not even needed 
as a variable name. The function argument of `nlm` works perfectly well with 
no named arguments! 

In this post I will introduce the L-function, which allows you to write the 
previous function call like this:

```r
nlm(f = L(-mean(dgamma(x, shape = ?1, rate = ?2, log = TRUE))), 
    p = c(1, 1))
```

The function created by L doesn't have any named arguments, but you can refer to 
the unnamed arguments by the question mark. Actually, the function `f` above is
equivalent to the following, just with some syntactic sugar on top.

```r
function(...) -mean(dgamma(x, ...[[1]], ...[[2]], log = TRUE)))
```

This works in the `nlm` case since whenever `f` has the signature `...` and 
receives a vector, it interprets this as `do.call` over the vector:

```{r, eval = TRUE}
f = function(...) ...[[1]]
f(c(1, 2, 3))
``` 

The consequence is that functions that depend on vectorization (e.g. 
`integrate`) won't work together with L.

### The L function
```{r, eval = TRUE}

call_replace = function(call) {
  if(length(call) > 1) {
    if(call[[1]] == quote(`?`)) 
      if(is.numeric(call[[2]])) 
        return(parse(text = paste0("...[[", eval(call[[2]]), "]]"))[[1]])
    
    new = as.call(lapply(1:length(call), function(i) call_replace(call[[i]])))
    names(new) = names(call)
    new
  } else call
}

L = function(call, quoted = FALSE) {
  call = if(!quoted) substitute(call) else call
  f = function(...) NULL
  body(f) = call_replace(call)
  environment(f) = parent.frame()
  f
}

```

What does it do? First, `call_replace` travels recursively through the call and 
replaces any instance of `?n` with `...[[n]]`. The `L` constructs a function 
from the modified call.

### Using magrittr
It's also possible to make a piped version of this:

```{r, eval = TRUE}
library("magrittr")

`%L>%` = function(lhs, rhs) {

  lhs_call = call(name = "function", 
                  quote(function(...) {})[[2]], 
                  call_replace(substitute(lhs)))
  
  rhs_call = substitute(rhs)
  
  eval(call("%>%", lhs_call, rhs_call))
}

-mean(dgamma(x, shape = ?1, rate = ?2, log = TRUE)) %L>% 
  nlm(p = c(1, 1)) %$%
  estimate
```

This piped version won't work if it is placed in the middle of a magrittr pipe. 
Take this example:

```{r, eval = TRUE, error = TRUE}
rgamma(100, 2, 7) %>%
  -mean(dgamma(shape = ?1, rate = ?2, log = TRUE)) %L>% 
  nlm(p = c(1, 1))
```

### Ideas to modify L

We could use `!n` to reference the $n$th argument of the created function. 
Moreover, they could be named and given default values. The propotype
`L(f(a = !1 == b <- c, 1, 2, 3))` will return a function of one argument named 
`b` with default value `c`.
